A Model to Predict Survival in Patients with End-Stage Liver Disease

### 2021-10-19 analysis/article.Rmd
feat: describe zlog in more detail

## Machine learning

@@

All statistical and machine learning analyses were performed using R version
`r paste0(sessionInfo()$R.version[c("major", "minor")], collapse = ".")`
[@R-base].
Prior to the analyses all laboratory values were zlog transformed
[@hoffmann2017; @R-zlog].
We compared `r n_lrns` different statistical and machine learning
algorithms using the mlr3 framework
[@mlr3; @R-mlr3; @mlr3proba2021; @R-mlr3proba].
The used algorithms are designed for analysis of survival data:
Cox proportional hazards regression model
[@cox1972; @survival-book; @R-survival],
penalized regressions,
namely two different implementations of the
lasso (least absolute shrinkage and selection operator) regression
[@tibshirani1997; @glmnet2011; @R-glmnet; @penalized2010; @R-penalized],
tow different implementations of the
ridge regression [@glmnet2011; @R-glmnet; @penalized2010; @R-penalized], and
the elastic net regression [@glmnet2011; @R-glmnet],
two different random forest implementations
[@ranger2017; @R-ranger; @randomForestSRC2008; @R-randomForestSRC],
extreme gradient boosting [@chen2016; @R-xgboost],
two different support vector machines [@vanbelle2011; @R-survivalsvm]
and a neural network [@katzman2018; @R-survivalmodels].

To choose the best model we used nested resampling to avoid feature- and
model-selection-bias.
Therefore, we applied a `r crossval$inner$param_set$values$folds`-fold inner
and a `r crossval$outer$param_set$values$folds`-fold outer cross-validation
that was repeated `r crossval$outer$param_set$values$repeats` times.
Subsequently we selected the model with the
highest median concordance index [@harrell1982].

>>>>

## Data processing and machine learning

@@

All data processing, statistical and machine learning analyses were performed
using R version
`r paste0(sessionInfo()$R.version[c("major", "minor")], collapse = ".")`
[@R-base].
Prior to the analyses all laboratory values were *zlog* transformed
as described in @hoffmann2017 and implemented in @R-zlog.

$$
z = (\log(x) - \frac{\log(LL) + \log(UL)}{2}) * \frac{3.92}{\log(UL) - \log(LL)}
$$

Where $LL$ and $UL$ are the lower and upper reference limit of
the corresponding laboratory diagnostic, respectively.
Missing laboratory measurements were replaced with a *zlog* value of zero.
In similar manner if information about complications or
comorbidities were missing we treat them as not present.

We compared `r n_lrns` different statistical and machine learning
algorithms using the mlr3 framework
[@mlr3; @R-mlr3; @mlr3proba2021; @R-mlr3proba].
The used algorithms are designed for analysis of survival data:
Cox proportional hazards regression model
[@cox1972; @survival-book; @R-survival],
penalized regressions,
namely two different implementations of the
lasso (least absolute shrinkage and selection operator) regression
[@tibshirani1997; @glmnet2011; @R-glmnet; @penalized2010; @R-penalized],
tow different implementations of the
ridge regression [@glmnet2011; @R-glmnet; @penalized2010; @R-penalized], and
the elastic net regression [@glmnet2011; @R-glmnet],
two different random forest implementations
[@ranger2017; @R-ranger; @randomForestSRC2008; @R-randomForestSRC],
extreme gradient boosting [@chen2016; @R-xgboost],
two different support vector machines [@vanbelle2011; @R-survivalsvm]
and a neural network [@katzman2018; @R-survivalmodels].

To choose the best model we used nested resampling to avoid feature- and
model-selection-bias.
Therefore, we applied a `r crossval$inner$param_set$values$folds`-fold inner
and a `r crossval$outer$param_set$values$folds`-fold outer cross-validation
that was repeated `r crossval$outer$param_set$values$repeats` times.
Subsequently we selected the model with the
highest median concordance index [@harrell1982]. If multiple models were
comparable we chose the simpler (more penalized) one.
